The Evolution of Phenotypically Invariant Genetic Networks

### 2017-08-09 kalman_walk_paper.tex
edited and replied to some comments.

% for comments in margins or not
\newif\ifmargincomments
\margincommentstrue
% \margincommentstrue
\ifmargincomments
    \usepackage{todonotes}
    \usepackage[left=1cm,right=6.5cm,top=3cm,bottom=3cm,nohead,nofoot,marginparwidth=6cm]{geometry}
    \newcommand{\plr}[1]{\todo[color=blue!25]{#1}}
    % inline comment version
    \newcommand{\plri}[1]{{\color{blue}\it #1}}
\else
    \newcommand{\plr}[1]{{\color{blue}\it #1}}
    \newcommand{\plri}[1]{\plr{#1}}
\fi

@@

%\begin{multicols}{2}

\begin{abstract}
Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks. The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes. Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set. We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit. We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
\plr{Saying "stalled" reflects negatively on other fields - rephrase?}
There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
\plr{I don't think the HW example will speak to our audience.  Also, better to say why math is good rather than why not-math is bad.}
However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}. 
This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
This outlook necessitates the advancement of conceptual frameworks of such precision, 
only mathematics will suffice, as models allow the development of concrete numerical predictions. 
Previously it has been suggested that any idealized study of evolution is incomplete 
without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}. 
\plr{probably some more recent Wagner papers in this line as well?}
%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

@@

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
\plr{not clear what you mean by "actually"}
Models with different parameter schemes, or even different mechanics can be equally accurate, 
but still not \emph{actually} agree with what is being modelled. 
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
In chemical engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.  
\plr{here you are jumping to evolution -- haven't said yet why evolution "can explore" these -- needs to say explicitly what we mean by neutral here}
Although this may frustrate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; 
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

In this paper we outline a theoretical framework to study the evolution of biological systems. 
\plr{rather than "we expect" maybe say that the framework could be applied to non-neutral evolution}
We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), 
however presently, we focus solely on neutral evolution, 
\plr{here's the definition of neutral, but it's not quite right?}
that is where phenotype is conserved over evolutionary time.

\plr{We don't "derive" it, we point out that Kalman already derived it.}
We derive an analytical description of the set of all linear biological systems with identical phenotypes 
\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".}
-- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. 
In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. 
Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. 
We define an organism's ``phenotype'' to be the time dynamics of molecular concentrations directly relevant to survival, 
\plr{tried to say it more precisely. better phrase than "component of phenotype"?}
and denote by $\phi_k(t)$ the concentration of the $k^\text{th}$ component of phenotype at time $t$.

@@

and an environmental input $u(t)$.
\plr{say what $u$ is (a vector)}
\plr{say in words what $\kappa$ is before the equation.}

@@

Where the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}. As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the systems \emph{external dynamics}.
The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment. 
\plr{Could say instead of "filters" that e.g. $C_{ij}$ is the amount by which the $j$th input $u_j$ affects the production of the $i$th transcription factor concentration.}
$C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.
\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}
Thus the phenotype is simply a convolution of the system organization and the environment, 
\plr{You've missed out $\kappa(0)$. Also, you can't have $t$ be both a limit of the integral and the variable being integrated over -- check this and get it right.}
    \phi(t) = \int_{0}^{t} C e^{A t} B u(t) dt ,
where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system.

\plr{How about expanding this minor comment into a short paragraph with more interpretation of not just $A$ but also $B$ and $C$? Should say why. 
Include talking through why $A_{ij}>0$ means that $j$ "upregulates" $i$.}
As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.

%We will now lay out the model in more general terms.
%Suppose that the \emph{internal state} of the system
%is parameterized by the concentrations of a collection of $n$ molecular species,
%$S_1, \ldots, S_n$,
%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
%There are also $m$ ``input'' species, whose concentrations are determined
%exogenously to the system,
%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
%and $\ell$ ``output'' species, whose concentrations are denoted
%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
%The output is merely a linear function of the internal state:
%\begin{align*}
%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
%\end{align*}
%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
%as it is ``hidden'' from direct selection.
%The rate at which the $i^\text{th}$ species is produced
%is a weighted sum of the concentrations of the other species
%as well as the input:
%\begin{align*}
%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
%\end{align*}
%In matrix notation, this is written more concisely as
%\begin{align} \label{eqn:lti_system}
%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
%    \phi(t) &= C \kappa(t) .
%\end{align}

  \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}

  \section*{Linear Evolutionary Systems}
\plr{instead title this section "Systems drift for linear systems"?}

\plr{This paragraph feels like the introduction.}
The literature is filled with detailed observations of molecular systems and their diversity. 
\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
There are examples of significant diversity in the networks underlying processes such as 
circadian rhythm \citep{sancar2008intelligent}, 
cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
pattern formation \plr{cite?}, 
and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level. 
How many different mechanisms have the same function? 
We urge the reader to consider the metaphorical black box.

Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics). 
Any linear and minimal system (a gene network) 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes 
-- has an identical phenotype up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= \bar{C} e^{\bar{A} t} \bar{B}
  \end{align}
\plr{what are we calling these -- not "biological system" -- that's too broad}
Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
    i.e. behaviour across many possible inputs.}
Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
  \begin{align}
    VB &= B \implies \bar{B} = B \\
    CV &= C \implies \bar{C} = C
  \end{align}
\plr{this sentence lacks an object (or some sentence part?)}
In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.

\plr{I don't think it makes sense to write things in general as a function of $\tau$,
    since in general the set of "free parameters" is complicated and we don't actually specify it.
    Better, maybe define $\mathcal{V}$ to be the set of all $V$ that preserve $B$ and $C$, then define $A(V)$ instead of $A(\tau)$...}
Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
and a \emph{Linear Evolutionary System} is, 
\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
  \begin{align}
    \mathcal{S}(\tau) := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
\plr{again, lacking an object}
That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.

More generally, \plr{introduce Kalman},
we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
  \calA_n(A_0) 
      &= \left\{
          A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
          A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

\plr{should say more about this decomposition here.}
Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$. This set can be completely parameterized using the \emph{Kalman Decomposition}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined. Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.

  \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks]

\plr{this is same as above, so reference instead of writing out again?}
    Let
    \begin{align*}
      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
    \end{align*}
    with $V(\tau)$ preserving both $B$ and $C$, then
    the set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in Example \ref{ex:oscillator} are given by
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\plri{Missing: a statement about when there are neutral directions, and how many.}

\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}

>>>>

% for comments in margins or not
\newif\ifmargincomments
\margincommentstrue
% \margincommentstrue
\ifmargincomments
    \usepackage{todonotes}
    \usepackage[left=1cm,right=6.5cm,top=3cm,bottom=3cm,nohead,nofoot,marginparwidth=6cm]{geometry}
    \newcommand{\plr}[1]{\todo[color=blue!25]{#1}}
    \newcommand{\js}[1]{\todo[color=green!25]{#1}}
    % inline comment version
    \newcommand{\plri}[1]{{\color{blue}\it #1}}
\else
    \newcommand{\plr}[1]{{\color{blue}\it #1}}
    \newcommand{\plri}[1]{\plr{#1}}
\fi

@@

\begin{abstract}
We introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks. The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes. Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set. We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit. We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%
%\begin{multicols}{2}
\section*{Introduction}

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
  \plr{Saying "stalled" reflects negatively on other fields - rephrase?} \js{I don't think it reflects negatively. Even if it did, as long as it's a true statement it should be fine.}.
There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
  \plr{I don't think the HW example will speak to our audience.  Also, better to say why math is good rather than why not-math is bad.} \js{deleted HW. I don't think we need to write an anodyne -- the current sentence is true enough.}
However, verbal theories are often insufficient, if not downright misleading \citep{servedio2014not}. 
This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
This outlook necessitates the advancement of conceptual frameworks of such precision, 
only mathematics will suffice, as models allow the development of concrete numerical predictions. 
%Previously it has been suggested that any idealized study of evolution is incomplete 
%without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.
\plr{probably some more recent Wagner papers in this line as well?}
%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

@@

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
%\plr{not clear what you mean by "actually"}\js{I think we discussed it's clear enough for an intro?}
Models with different parameter schemes, or even different mechanics can be equally accurate, 
but still not \emph{actually} agree with what is being modelled. 
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
In chemical engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.  
\plr{here you are jumping to evolution -- haven't said yet why evolution "can explore" these -- needs to say explicitly what we mean by neutral here}
Although this may frustrate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; 
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

In this paper we outline a theoretical framework to study the evolution of biological systems. 
  \plr{rather than "we expect" maybe say that the framework could be applied to non-neutral evolution}\js{fixed}
Presently, we focus solely on a neutral scenario, 
that is where phenotype is conserved over evolutionary time. 
however, this framework could be applied to a wider set of evolutionary scenarios. 
\plr{here's the definition of neutral, but it's not quite right?} \js{how's that?}

We present an analytical description of the set of all linear biological systems with identical phenotypes 
\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".} \js{added the word ``linear.''}
-- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all linear biological systems can, in principal, can undergo systems drift. 
In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. 
Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

\js{Will rewrite this section again.} Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. 
We define an organism's ``phenotype'' to be the temporal dynamics of molecular concentrations directly relevant to survival, 
and denote by $\phi_k(t)$ the concentration of the $k^\text{th}$ element of the phenotype at time $t$, given an input; for instance $\phi_k(t)$ could describe the concentrations of a particular enzyme.

@@

and an environmental input vector $u(t)$, and the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}. 

@@

As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the system's \emph{external dynamics}.
The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the system's current state (the kryptotype), the system architecture $(A,B,C)$ -- such as the wiring of a gene regulatory network, and the environmental input $u(t)$ to the system. The architecture of $A$ determines exactly how the internal molecular concentrations regulate one another's expression as $A_{ij}$ is the magnitude at which transcription factor $\kappa_j(t)$ regulates transcription factor $\kappa_i(t)$. If $A_{ij} > 0$, we say that $\kappa_j$ upregulate $\kappa_i$, and if $A_{ij} < 0$, we say that $\kappa_j$ down-regulates $\kappa_i$. Furthermore, we interpret the $i^\text{th}$ row of $A$ to be the \emph{cis}-regulatory element (or promoter) for gene $i$.  
The architecture of $B$ determines how the system processes the environmental intput, and the structure of $C$ converts the internal dynamics to external dynamics. That is if $\mathcal{S}$ is a metabolic system, $C_{ij}$ is the amount at which the $j$th input metabolite $u_j$ affects the production of the $i$th enzyme concentration.

\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}
Thus the phenotype is simply a convolution of the system organization and the environment, 
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
where we refer to $h(t) := Ce^{A t}B$ as the system's \emph{impulse response}. 
%We will now lay out the model in more general terms.
%Suppose that the \emph{internal state} of the system
%is parameterized by the concentrations of a collection of $n$ molecular species,
%$S_1, \ldots, S_n$,
%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
%There are also $m$ ``input'' species, whose concentrations are determined
%exogenously to the system,
%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
%and $\ell$ ``output'' species, whose concentrations are denoted
%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
%The output is merely a linear function of the internal state:
%\begin{align*}
%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
%\end{align*}
%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
%as it is ``hidden'' from direct selection.
%The rate at which the $i^\text{th}$ species is produced
%is a weighted sum of the concentrations of the other species
%as well as the input:
%\begin{align*}
%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
%\end{align*}
%In matrix notation, this is written more concisely as
%\begin{align} \label{eqn:lti_system}
%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
%    \phi(t) &= C \kappa(t) .
%\end{align}

  \section*{Gene Network Drift}

%\plr{This paragraph feels like the introduction.}
%The literature is filled with detailed observations of molecular systems and their diversity. 
%\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
%There are examples of significant diversity in the networks underlying processes such as 
%circadian rhythm \citep{sancar2008intelligent}, 
%cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
%pattern formation \plr{cite?}, 
%and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
Despite a symmetry in functionality or phenotype systems can often differ, sometimes substantially, at the molecular level. 
How many different mechanisms have the same function?

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal system (a gene network) 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes 
-- has an identical impulse response and therefore an identical phenotype given an identical input $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
\plr{what are we calling these -- not "biological system" -- that's too broad}\js{how about ``molecular systems''?}
Two molecular systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

%Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
%That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
%\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
%    i.e. behaviour across many possible inputs.}
%Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
%  \begin{align}
%    VB &= B \implies \bar{B} = B \\
%    CV &= C \implies \bar{C} = C
%  \end{align}
%\plr{this sentence lacks an object (or some sentence part?)}
%In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore, if a system is minimal, the set of all phenotypically identical systems $\mathcal{S}(V)$ can be parameterized as,
    \begin{align}
      \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  \end{align}
  where $V$ is any invertible matrix and it's elements are free parameters. 
More generally,
we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
  \calA_n(A_0) 
      &= \left\{
          A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
          A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

\plr{should say more about this decomposition here.}
Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$. This set can be completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined. Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection*{Speciation}

%Speciation, or the formation of new species, is often a consequence of reprdouctive incompatibility among populations -- usually via hybrid sterility or inviability. To discuss speciaiton we first need to describe how reproduction is modelled.
 %A diploid first generation hybird's ($F_1$'s) genetic regulation is a consequence of its two genomes, as a copy of genes and regulatory sequences from each parent is present.
%\plr{should have a stronger justification for the assertion that $F_1$ is the arithmetic average of the parents than "Thus we say"}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks. 
%\plr{I don't think we say that $A$ is a "gene network architecture".}
%Therefore the phenotypic dynamics typical of an $F_{1}$ cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid crosses $F_2$ between allopatric populations, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.

%\begin{align}
%    F_{1}(\tau, \hat{\tau}) &= \frac{ A(\tau) + A(\hat{\tau}) }{2}  .
%   h(t, F_{1}) &= C e^{\left( F_{1}\left( \tau, \hat{\tau} \right) t \right)} B
%\end{align}
%\begin{align*}
% G\left(r, \tau, \hat{\tau}\right) = Q(r) A( \tau ) + \left( I - Q(r) \right) A( \hat{\tau} )
%\end{align*}
%\begin{align}
%    F_{2}(i, j) = \frac{G (i) + G(j)}{2}
%\end{align}

%Fitness $\fit$ is a function (usually a Guassian) of an organism's weighted phenotypic distance from an optimum,
%  $\fit\left( \widehat{\phi}(t) \right) = \exp \left(- \int_{0}^{t} \frac{\rho(s)}{\beta} \left\lVert \phi(s) - \widehat{\phi}(s) \right\rVert^{2} ds  \right)$
%where $\rho(t)$ is some weighting function and $\beta$ is a parameter.

%We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
%\begin{align*}
%      \Incompat = \frac{2 \left\langle \fit \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \fit \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \fit \left(\phi_{P_{2}}\right) \right\rangle} ,
%\end{align*}
%where angled brackets imply averaging.  \plr{averaging over what? And, this is as others define - cite.}
